---
title: "Assignment 2, Task 1"
author: "Alessandra Puig-Santana"
date: "2023-02-13"
output: html_document
---

## Task 1 - Palmetto binary logistic regression (individual)

In a separate .Rmd, using the Florida palmetto data (palmetto.csv), use binary logistic regression to test feasibility of using variables plant height (height), canopy length (length), canopy width (width), and number of green leaves (green_lvs) to classify whether a palmetto is species Serenoa repens or Sabal etonia.

**Data source:** Abrahamson, W.G. 2019. Survival, growth and biomass estimates of two dominant palmetto species of south-central Florida from 1981 - 2017, ongoing at 5-year intervals ver 1. Environmental Data Initiative. <https://doi.org/10.6073/pasta/f2f96ec76fbbd4b9db431c79a770c4d5>

```{r setup, echo = TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Set up chunk

library(tidyverse)
library(here)
library(kableExtra)
library(AICcmodavg)
library(GGally)
library(tidymodels)
library(broom)
library(janitor)
library(patchwork)

## Read in data
palmetto <- read_csv(here('palmetto.csv'))

```

### Data exploration

Explore differences in height, canopy length, canopy width, and green leaves for the two species. If you prefer, combine the figures into a compound figure using {patchwork} or {cowplot}. Below your data visualizations, add a sentence or two with a takeaway from the plots, e.g., based on these plots, which predictor variables are more likely to help classify species correctly?

#### Setting up data to explore

```{r}
# Species_id 1 - Serenoa repens
# Species_id 2 - Sabal etonia

palmetto_sub <- palmetto %>% 
  select(species, height:green_lvs) %>% 
  mutate(species = case_when(species == "1" ~ "Serenoa repens",
                   species == "2" ~ "Sabal etonia")) %>%
  drop_na()

palmetto_sub$species <- factor(palmetto_sub$species)

class(palmetto_sub$species) # check to make sure it worked
levels(palmetto_sub$species) #see which one is 0 and which is 1

# visualize using ggpairs
# ggpairs(data = palmetto_sub, aes(color = species))

```

### Data visualizations 
#### Plot set 1 - Height and canopy length 
```{r}
p1 <- ggplot(data = palmetto_sub, 
             aes(x = height, y = length)) +
  geom_point(aes(color = species)) +
  labs(x = 'Height (cm)',
       y = 'Canopy length (cm)') + 
  scale_color_manual(values = c('peachpuff', 'salmon')) +
  theme_minimal() +
  theme(
    #customize legend elements
    legend.position = c(0.12, 0.8),
    legend.title = element_text(face = 'bold', size = 7),
    legend.text = element_text(face = 'italic', size = 7),
    legend.background = element_rect(fill = 'white'),
    legend.key = element_rect(fill = 'white'),
    legend.key.size = unit(1, 'line'),
    #customize axis elements
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    axis.title.x = element_text(size = 10, face = 'bold'),
    axis.title.y = element_text(size = 10, face = 'bold'))

p1

p2 <- ggplot(data = palmetto_sub, aes(x = height, y = length)) +
  geom_point(aes(color = species), alpha = 0.8, show.legend = FALSE) +
  labs(x = 'Height (cm)',
       y = 'Canopy length (cm)') +
  facet_wrap(~species) +
  scale_color_manual(values = c('peachpuff', 'salmon'))+
  theme_minimal()+
  theme(
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    axis.title.x = element_text(size = 9, face = 'bold'),
    axis.title.y = element_text(size = 9, face = 'bold'),
    panel.background = element_rect(fill = 'white'))

p2

# Combining both graphs
p1 + p2
```
#### Plot set 2 - Height vs Canopy width 
```{r}
p3 <- ggplot(data = palmetto_sub, 
             aes(x = height, y = width)) +
  geom_point(aes(color = species)) +
  labs(x = 'Height (cm)',
       y = 'Canopy width (cm)') +
  scale_color_manual(values = c('palegreen', 'palegreen4')) +
  theme_minimal() +
  theme(
    #customize legend elements
    legend.position = c(0.12, 0.8),
    legend.title = element_text(face = 'bold', size = 7),
    legend.text = element_text(face = 'italic', size = 7),
    legend.background = element_rect(fill = 'white'),
    legend.key = element_rect(fill = 'white'),
    legend.key.size = unit(1, 'line'),
    #customize axis elements
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    axis.title.x = element_text(size = 10, face = 'bold'),
    axis.title.y = element_text(size = 10, face = 'bold'))

p3

p4 <- ggplot(data = palmetto_sub, aes(x = height, y = width)) +
  geom_point(aes(color = species), alpha = 0.8, show.legend = FALSE) +
  labs(x = 'Height (cm)',
       y = 'Canopy length (cm)') +
  facet_wrap(~species) +
  scale_color_manual(values = c('palegreen', 'palegreen4'))+
  theme_minimal()+
  theme(
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    axis.title.x = element_text(size = 9, face = 'bold'),
    axis.title.y = element_text(size = 9, face = 'bold'),
    panel.background = element_rect(fill = 'white'))

# Combining graphs 
p3 + p4
```

#### Plot 4 - Green leaves count vs Count per species 
```{r}
p4 <- ggplot(data = palmetto_sub, aes(x = green_lvs, fill = species)) +
  geom_histogram() +
  facet_wrap(~species)+
  theme_minimal()+
  labs( x = "Number of Green Leaves",
        y = "Count per species")+
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
        plot.subtitle = element_text(hjust = 0.5, face = "bold", size = 11),
        axis.title.x = element_text(vjust = 0.2, face = "bold", size = 10),
        axis.title.y = element_text(face = "bold", size = 10),
        axis.text = element_text(size = 8),
        plot.background = element_rect(fill = "white"),
        legend.position = "none")+
  scale_fill_manual(values = c("darkolivegreen1", "forestgreen"))

p4
```

### Binary Logistic Regression
```{r}
# Log odds of plant type using plant height, canopy length, canopy width and green leaves as predictor variable

# Formula 1 
f1 = species ~ height + length + width + green_lvs

# BLR of f1
blr1 <- glm(formula = f1, 
            data = palmetto_sub, 
            family = "binomial")

# Log odds of plant type using plant height, canopy width and green leaves (i.e., drop canopy length for this model)

# Formula 2
f2 = species ~ height + width + green_lvs

# BLR of f2
blr2 <- glm(formula = f2, 
            data = palmetto_sub, 
            family = "binomial")
```

```{r}
# Make it tidy
blr1_tidy <- tidy(blr1)
blr2_tidy <- tidy(blr2)

# Tables with log odds
blr1_tidy %>% 
  kable(caption = "**Table 1.** BLR results for Model 1.",
        col.names = c("Variable", "Coefficient", "Standard error", "Statistic", "p-value")) %>% 
  kable_styling(full_width = FALSE,
                bootstrap_options = "hover",
                position = "center")
blr2_tidy %>% 
  kable(caption = "**Table 2.** BLR results for Model 2.",
        col.names = c("Variable", "Coefficient", "Standard error", "Statistic", "p-value")) %>% 
  kable_styling(full_width = FALSE,
                bootstrap_options = "hover",
                position = "center")
```

### Model testing 
```{r}
# AIC 
aictab(list(palmetto_blr1, palmetto_blr2)) %>% 
    kable(col.names = c("Model", "Parameters", "AIC", "Delta AIC", "Model Likelihood", "AIC Weight", "Log Likelihood", "Cumulative Weight"),
        caption = "Table 3: AIC Results") %>% 
  kable_classic(position = "center")

# BIC 
bictab(list(palmetto_blr1, palmetto_blr2)) %>% 
    kable(col.names = c("Model", "Parameters", "BIC", "Delta BIC", "Model Likelihood", "BIC Weight", "Log Likelihood", "Cumulative Weight"),
        caption = "Table 4: BIC Results") %>% 
  kable_classic(position = "center")
```

```{r}
# Cross fold validation

## set seed for reproducibility 
set.seed(123)

## set number of folds to 10 and repeat it 5 times
tidy_folds <- vfold_cv(palmetto_sub, v = 10, repeats = 10)

## Create general BLR model
blr_model <- logistic_reg() %>% 
  set_engine('glm')

# Workflow that bundles the logistic model and formula 1
blr_tidy_wf1 <- workflow() %>%
  add_model(blr_model) %>%
  add_formula(f1)

blr_tidy_cv_f1 <- blr_tidy_wf1 %>%
  fit_resamples(tidy_folds)

# Use functions from the tune package to extract metrics
metrics1<- collect_metrics(blr_tidy_cv_f1)

# make a nice table to display model 1 metrics from CV
metrics <- metrics1 %>% 
  kable(col.names = c("Metric",
                      "Estimator",
                      "Mean",
                      "n",
                      "Standard Error",
                      "Configuration"),
        caption = "Table 5: Cross Validation Metrics Model 1") %>% 
  kable_classic(position = "center")

metrics
```

```{r}
# Model 2
blr_tidy_wf2 <- workflow() %>%
  add_model(blr_model) %>%
  add_formula(f2)

blr_tidy_cv_f2 <- blr_tidy_wf2 %>%
  fit_resamples(tidy_folds)

# Use functions from the tune package to extract metrics
metrics2 <- collect_metrics(blr_tidy_cv_f2)

metrics_2 <- metrics2 %>% 
  kable(col.names = c("Metric", "Estimator", "Mean", "n", "Standard Error", "Configuration"),
        caption = "Table 6: Cross Validation Metrics Model 2") %>% 
  kable_classic(position = "center")

metrics_2
```

